---
title: "CD8+ HS Dataset analysis and Integration"
author: "Veronika Niederlova and Juraj Michalik"
date: "4/26/2021"
output: rmdformats::material
---

This script pre-processes one of data-sets used to build HS Atlas presented, among others, in manuscript written by Tsyklauri et al., 2022.
The license applicable to this script (as well as others in this directory) can be consulted in GitHub repository (MIT, see respective file for details).

Some of data-sets used in this atlas are downloaded from 10X site and as such are subject to their license;
please see the list of used data-sets and consult their license (Creative Commons License).

# Initialization 

Analysis Donor 3 data-set.

First we load all necessary packages here.

```{r libs, include=FALSE}
knitr::opts_chunk$set(include = TRUE, warning = FALSE, message = FALSE, 
                      error = TRUE, cache = TRUE, cache.lazy = FALSE)
library(Seurat, lib.loc = '/usr/lib64/R/library')
library(future)
library(ggplot2)
library(dplyr)
library(tidyr)
library(cowplot)
library(gridExtra)
library(dbscan)
library(scatterpie)
library(reshape2)
library(ggrepel)
library(scds)
library(SingleCellExperiment)

# negation of %in% 
`%nin%` = Negate(`%in%`)
```

## R Markdown

```{r, message = F}
donor3tab <- Read10X('Initial_Data/Donor3/filtered_feature_bc_matrix/')
donor3tab$`Gene Expression` <- donor3tab$`Gene Expression`[!grepl('^TR(A|B)(V|D|J)', rownames(donor3tab$`Gene Expression`)),] 
```

We create Seurat object with Antibody assay. We also remove raw count matrices as we no longer need them and here the space might be an issue.

```{r}
# create seurat object
donor3 <- CreateSeuratObject(donor3tab$`Gene Expression`, 'Donor3_HS_CD8',  min.cells = 3, min.features = 200)
ab_assay <- donor3tab$`Antibody Capture`[,colnames(donor3tab$`Antibody Capture`) %in% 
                                          colnames(donor3)]
ab_assay <- CreateAssayObject(counts = ab_assay)
donor3[['Antibodies']] <- ab_assay
remove(donor3tab, ab_assay)
```

Perform QC.

```{r}
donor3[["percent.mt"]] <- PercentageFeatureSet(object = donor3, pattern = "^MT-")
donor3[["percent.rp"]] <- PercentageFeatureSet(object = donor3, pattern = "^RP[S|L]")
print(VlnPlot(donor3, features = c("nFeature_RNA", "nCount_RNA", "percent.mt", "percent.rp"), ncol = 4, pt.size = 0))
```

Filter out low QC cells and VDJ doublets. Here it looks better to keep cells even with high read counts because they may not be doublets but some cells with high transcription activity. We follow by adding VDJ data.

```{r}
vdj.data <- load.10X.VDJ('Initial_Data/vdj_v1_hs_aggregated_donor3_all_contig_annotations.csv')

donor3 <- subset(donor3, subset = nFeature_RNA > 200)
vdj.doublets <- trace.VDJ.duplicates(donor3, vdj.data)
donor3 <- subset(donor3, cells = colnames(donor3)[!(colnames(donor3) %in% vdj.doublets)])
donor3 <- create.VDJ.metadata(donor3, vdj.data, filter.vdj = T, generate.clonotypes = F)
```

# First Analysis

Now we continue with the analysis.

```{r, results = F, include=F}
donor3 <- NormalizeData(object = donor3)
donor3 <- FindVariableFeatures(donor3, selection.method = "vst", nfeatures = 1000, 
                                          verbose = FALSE)
  
donor3 <- ScaleData(donor3, verbose = FALSE)
donor3 <- RunPCA(donor3, npcs = 20, verbose = FALSE)
donor3 <- RunUMAP(donor3, reduction = "pca", dims = 1:20)
  
donor3 <- FindNeighbors(donor3, dims = 1:20)
donor3 <- FindClusters(donor3, resolution = 0.5)
```

```{r}
print(DimPlot(donor3, reduction = "umap", group.by = "seurat_clusters", label = TRUE))
print(VlnPlot(donor3, features = c("nFeature_RNA", "nCount_RNA", "percent.mt", "percent.rp"), ncol = 2, pt.size = 0))
```

```{r, fig.width = 10, fig.height = 15}
print(FeaturePlot(donor3, c('CD8A', 'CD8B', 'CCR7', 'CD62', 'CD28', 'CD27', 'CD44'), min.cutoff = 0))
print(FeaturePlot(donor3, c('CXCR3', 'CX3CR1', 'TBX21', 'KLRB1', 'IL2RB', 'IL7R'), min.cutoff = 0))
```

# Filtering and subsequent analysis

## First filtering

Clusters 0, 1, 6, 11 and 12 are clusters with plenty of mitochondrial genes and low gene count - we'll remove them. Cluster 15 has most likely no CD8 cells so we remove it as well. We also remove all cells with more than 10% mitochondrial genes in this step. Cluster 16 has rather strange Violin plots and considering its relatively low cell number, we decided to remove it. 

```{r, results = F, include = F}
donor3 <- subset(donor3, subset =  percent.mt < 10 & seurat_clusters %nin% c(0,1,6,11,12,15,16))

donor3 <- NormalizeData(object = donor3)
donor3 <- FindVariableFeatures(donor3, selection.method = "vst", nfeatures = 1000, 
                                          verbose = FALSE)
  
donor3 <- ScaleData(donor3, verbose = FALSE)
donor3 <- RunPCA(donor3, npcs = 20, verbose = FALSE)
donor3 <- RunUMAP(donor3, reduction = "pca", dims = 1:20)
  
donor3 <- FindNeighbors(donor3, dims = 1:20)
donor3 <- FindClusters(donor3, resolution = 0.5)
```

And repeat the visualization.

```{r}
print(DimPlot(donor3, reduction = "umap", group.by = "seurat_clusters", label = TRUE))
print(VlnPlot(donor3, features = c("nFeature_RNA", "nCount_RNA", "percent.mt", "percent.rp"), ncol = 2, pt.size = 0))
```

```{r, fig.width = 10, fig.height = 15}
print(FeaturePlot(donor3, c('CD8A', 'CD8B', 'CCR7', 'CD62', 'CD28', 'CD27', 'CD44'), min.cutoff = 0))
print(FeaturePlot(donor3, c('CXCR3', 'CX3CR1', 'TBX21', 'KLRB1', 'IL2RB', 'IL7R'), min.cutoff = 0))
```

## Second filtering

New cluster 10 is strange and contains very small number of cells (indicating an outlier), so we remove it.

```{r, results = F, include = F}
donor3 <- subset(donor3, subset =  percent.mt < 10 & seurat_clusters %nin% c(10))

donor3 <- NormalizeData(object = donor3)
donor3 <- FindVariableFeatures(donor3, selection.method = "vst", nfeatures = 1000, 
                                          verbose = FALSE)
  
donor3 <- ScaleData(donor3, verbose = FALSE)
donor3 <- RunPCA(donor3, npcs = 20, verbose = FALSE)
donor3 <- RunUMAP(donor3, reduction = "pca", dims = 1:20)
  
donor3 <- FindNeighbors(donor3, dims = 1:20)
donor3 <- FindClusters(donor3, resolution = 0.5)
```

```{r}
print(DimPlot(donor3, reduction = "umap", group.by = "seurat_clusters", label = TRUE))
print(VlnPlot(donor3, features = c("nFeature_RNA", "nCount_RNA", "percent.mt", "percent.rp"), ncol = 2, pt.size = 0))
```

```{r, fig.width = 10, fig.height = 15}
print(FeaturePlot(donor3, c('CD8A', 'CD8B', 'CCR7', 'CD62', 'CD28', 'CD27', 'CD44'), min.cutoff = 0))
print(FeaturePlot(donor3, c('CXCR3', 'CX3CR1', 'TBX21', 'KLRB1', 'IL2RB', 'IL7R'), min.cutoff = 0))
```

# Saving data

Looks good. Saving the data.

```{r, fig.width = 10, fig.height = 15}
saveRDS(donor3, paste0('Donors_Datasets/Donor_3_VDJ.rds'))
```